Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-20 main.tex
def of FOCP

\newcommand{\rectif}[1]{\widehat{#1}}
\newcommand{\graphof}[1]{\llbracket#1\rrbracket}
\newcommand{\compl}[1]{#1^\complement}
%\newcommand{\sublist}[1]{\set{#1}}
%\newcommand{\subst}[2]{#1\mathord{\scriptstyle\mapsto} #2}
\newcommand{\sublist}[1]{[#1]}
\newcommand{\subst}[2]{#1/#2}
\newcommand{\ssubst}[2]{\sublist{\subst{#1}{#2}}}
\newcommand{\substof}[1]{\sigma_{\!#1}}
\newcommand{\linkingof}[1]{\sim_{#1}}
\newcommand{\mapof}[1]{\lfloor{#1}\rfloor}

\newcommand{\form}[1]{\mathrm{fm}(#1)}
\newcommand{\formtree}[1]{\mathcal{F}(#1)}

@@

We start from a countable set $\VAR=\set{x,y,z,\ldots}$ of variables,
a countable set $\FUN=\set{f,g,\ldots}$ of function symbols, and a
countable set $\PRED=\set{p,q,\ldots}$ of predicate symbols. Each function symbol and
each predicate symbol has a finite arity. The grammars below generate
the set $\TERM$ of \bfit{terms}, denoted by $s,t,u,\ldots$, the set
$\ATOM$ of \bfit{atoms}, denoted by $a,b,c,\ldots$, and the set
$\FORM$ of \bfit{formulas}, denoted by $A,B,C,\ldots$:
\begin{equation*}
  \begin{array}{r@{~}l}
    t &\coloneqq~ x \mid f(t_1,\dots,t_n)
    \\[.5ex]
    a &\coloneqq~ \ttt\mid\fff\mid p(t_1,\dots,t_n) \mid \dual p(t_1,\dots,t_n)
    \\[.5ex]
    A &\coloneqq~ a \mid A\vlan A \mid A\vlor A \mid \exists x.A \mid \forall x.A
\end{array}
\end{equation*}
where the arity of $f$ and $p$ is $n$. Note, that in this paper we
consider the truth constants $\ttt$ (\emph{true}) and $\fff$
(\emph{false}) as atoms, and we consider all formulas in negation
normal form. The \bfit{negation} $(\dual\cdot)$ is defined for all
atoms and formulas via the De Morgan laws as follows:
\begin{equation*}
  \begin{array}{cc}
  \dual\ttt =\fff &
  \dual{p(t_1,\ldots,t_n)} = \dual{p}(t_1,\ldots,t_n) \\[.5ex]
  \dual\fff = \ttt &
  \dual{\dual p(t_1,\ldots,t_n)} = p(t_1,\ldots,t_n) \\[.5ex]
  \dual{\dual a} = a & \dual{\exists x.A} = \forall x.\dual{A} \\[.5ex]
  \dual{A \vlan B} = \dual{A} \ \vlor \ \dual{B} & \dual{\forall x.A} = \exists x.\dual{A} 
  \\[.5ex]
  \dual{A \vlor B} = \dual{A} \ \vlan \ \dual{B} &  
  \end{array}
\end{equation*}
\todo{add missing equations and make it look nice}
\lutz{don't we need   $\dual{\dual a} = a$? (then it should follow that $\dual{\dual A} = A$)}

\juihsuan{isn't it given by the first four equations? I think that we use the
notation of negation for atoms and formulas. We may also need to precise that each
predicate symbol $p$ has a dual $\dual{p}$.}\lutz{we either have to say $\dual{\dual a} = a$ or $\dual{\dual p} = p$. I prefer the first.}
\juihsuan{okay I see. I didn't realize that the negation is regarded as a
connective for predicate symbols. The notation got me confused since $\dual{p}$
is also a predicate symbol.}

@@

A \bfit{substitution} is a function $\sigma\colon\VAR\to\TERM$ that is
the identity almost everywhere. We denote substitutions as
$\sigma=\sublist{\subst{x_1}{t_1},\ldots,\subst{x_n}{t_n}}$, where
$\sigma(x_i)=t_i$ for $i=1..n$ and $\sigma(x)=x$ for all
$x\notin\set{x_1,\ldots,x_n}$. 
We write $A\sigma$ for the formula
obtained from $A$ by applying $\sigma$, i.e., by simultaneously
replacing all occurrences of $x_i$ by $t_i$.
A \bfit{renaming} is a substitution $\sigma$ with $\sigma(x)\in\VAR$ for all variable $x$.

@@

A graph $\gG$ is \bfit{labelled} in a set $L$ if each vertex $v\in\vG$
has an element $\ell(v)\in L$ associated with it, its \bfit{label}.  A
graph $\gG$ is (partially) \bfit{coloured} if it carries a partial
equivalence relation $\linkingof\gG$ on $\vG$; each equivalence class is a
\bfit{colour}.  A \bfit{vertex renaming} of $\gG=\tuple{\vG,\eG}$
along a bijection $(\hat\cdot)\colon\vG\to\hat\vG$ is the graph
$\hat\gG=\tuple{\hat\vG,\set{\hat v\hat w\mid vw\in\eG}}$, with
colouring and/or labelling inherited (i.e., $\hat v\samecol\hat w$ if
$v\samecol w$, and $\ell(\hat v)=\ell(v)$). Following standard graph
theory, we identify graphs modulo vertex renaming.

@@

@@

\subsection{Skew Bifibrations}
\lutz{$\phi$ instead of $f$}
A graph homomorphism $\phi\colon \tuple{\vG, \eG} \rightarrow \tuple{\vH, \eH}$ is a \bfit{fibration} if
for all $v \in \vG$ and $w\phi(v) \in \eH$, there exists a unique $\tilde{w} \in
\vG$ such that $\tilde{w}v \in \eG$ and $\phi(\tilde{w}) = w$, and is a \bfit{skew fibration}
if for all $v \in \vG$ and $w\phi(v) \in \eH$ there exists $\tilde{w} \in \vG$ such that
$\tilde{w}v \in \eG$ and $\phi(\tilde{w})w \notin \eH$.
A directed graph homomorphism is a \bfit{fibration} if \todo{}

\lutz{do we really need that?
A \bfit{fograph homomorphism}
is a graph homomorphism between the underlying graphs of the fographs
considered. A fograph homomorphism $f : \gG \rightarrow \gH$ is
\bfit{label-preserving} if for all $v \in \vG, label(v) = label(f(v))$, and is
\bfit{existential-preserving} if for all existential binder $b$ in $\gG$, the
vertex $f(b)$ is an existential binder in $\gH$.
}

\lutz{I think we need \bfit{fograph homomorphism} maps binders to binders and literals to literals, preserving the predicates.}

\begin{definition}
  Let $\gG$ and $\gH$ be fographs. A \bfit{skew bifibration} $\phi\colon\gG\to\gH$ is a fograph homomorphism that is a skew fibration on $\tuple{\vG, \eG} \to \tuple{\vH, \eH}$ and a fibration on the binding graphs $\bG\to\bH$.
  %etween fographs is a label-preserving and
%existential-preserving fograph homomorphism whose restriction to the
%binding graphs is a fibration.
\end{definition}

@@

>>>>

\newcommand{\rectif}[1]{\widehat{#1}}
\newcommand{\graphof}[1]{\llbracket#1\rrbracket}
\newcommand{\compl}[1]{#1^\complement}
%\newcommand{\sublist}[1]{\set{#1}}
%\newcommand{\subst}[2]{#1\mathord{\scriptstyle\mapsto} #2}
\newcommand{\sublist}[1]{[#1]}
\newcommand{\subst}[2]{#1/#2}
\newcommand{\ssubst}[2]{\sublist{\subst{#1}{#2}}}
\newcommand{\substof}[1]{\sigma_{\!#1}}
\newcommand{\linkingof}[1]{\sim_{#1}}
\newcommand{\mapof}[1]{\lfloor{#1}\rfloor}
\newcommand{\labelof}[1]{\ell(#1)}
\newcommand{\form}[1]{\mathrm{fm}(#1)}
\newcommand{\formtree}[1]{\mathcal{F}(#1)}

@@

We start from a countable set $\VAR=\set{x,y,z,\ldots}$ of variables,
a countable set $\FUN=\set{f,g,\ldots}$ of function symbols, and a
countable set $\PRED=\set{p,q,\ldots}$ of predicate symbols. Each function symbol and
each predicate symbol has a finite arity. The grammars below generate
the set $\TERM$ of \bfit{terms}, denoted by $s,t,u,\ldots$, the set
$\ATOM$ of \bfit{atoms}, denoted by $a,b,c,\ldots$, and the set
$\FORM$ of \bfit{formulas}, denoted by $A,B,C,\ldots$:
\begin{equation*}
  \begin{array}{r@{~}l}
    t &\coloneqq~ x \mid f(t_1,\dots,t_n)
    \\[.5ex]
    a &\coloneqq~ \ttt\mid\fff\mid p(t_1,\dots,t_n) \mid \dual p(t_1,\dots,t_n)
    \\[.5ex]
    A &\coloneqq~ a \mid A\vlan A \mid A\vlor A \mid \exists x.A \mid \forall x.A
\end{array}
\end{equation*}
where the arity of $f$ and $p$ is $n$. Note, that in this paper we
consider the truth constants $\ttt$ (\emph{true}) and $\fff$
(\emph{false}) as atoms, and we consider all formulas in negation
normal form. The \bfit{negation} $(\dual\cdot)$ is defined for all
atoms and formulas via the De Morgan laws as follows:
\begin{equation*}
  \begin{array}{ccc}
    \dual{\dual a} = a
    &
  \dual\ttt =\fff &
  \dual{p(t_1,\ldots,t_n)} = \dual{p}(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual\fff = \ttt &
  \dual{\dual p(t_1,\ldots,t_n)} = p(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual{\exists x.A} = \forall x.\dual{A}
  &
  \dual{A \vlan B} = \dual{A} \ \vlor \ \dual{B} 
  \\[.5ex]
  &
  \dual{\forall x.A} = \exists x.\dual{A}
  &
%  \\[.5ex]
  \dual{A \vlor B} = \dual{A} \ \vlan \ \dual{B}
  \end{array}
\end{equation*}

@@

A \bfit{substitution} is a function $\sigma\colon\VAR\to\TERM$ that is
the identity almost everywhere. We denote substitutions as
$\sigma=\sublist{\subst{x_1}{t_1},\ldots,\subst{x_n}{t_n}}$, where
$\sigma(x_i)=t_i$ for $i=1..n$ and $\sigma(x)=x$ for all
$x\notin\set{x_1,\ldots,x_n}$. 
We write $A\sigma$ for the formula
obtained from $A$ by applying $\sigma$, i.e., by simultaneously
replacing all occurrences of $x_i$ by $t_i$.
A \bfit{variable renaming} is a substitution $\sigma$ with $\sigma(x)\in\VAR$ for all variables $x$.

@@

A graph $\gG$ is \bfit{labelled} in a set $L$ if each vertex $v\in\vG$
has an element $\labelof v\in L$ associated with it, its \bfit{label}.  A
graph $\gG$ is (partially) \bfit{coloured} if it carries a partial
equivalence relation $\linkingof\gG$ on $\vG$; each equivalence class is a
\bfit{colour}.  A \bfit{vertex renaming} of $\gG=\tuple{\vG,\eG}$
along a bijection $(\hat\cdot)\colon\vG\to\hat\vG$ is the graph
$\hat\gG=\tuple{\hat\vG,\set{\hat v\hat w\mid vw\in\eG}}$, with
colouring and/or labelling inherited (i.e., $\hat v\samecol\hat w$ if
$v\samecol w$, and $\ell(\hat v)=\ell(v)$). Following standard graph
theory, we identify graphs modulo vertex renaming.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Skew Bifibrations}

A graph homomorphism $\phi\colon \tuple{\vG, \eG} \rightarrow
\tuple{\vH, \eH}$ is a \bfit{fibration} if for all $v \in \vG$ and
$w\phi(v) \in \eH$, there exists a unique $\tilde{w} \in \vG$ such
that $\tilde{w}v \in \eG$ and $\phi(\tilde{w}) = w$, and is a
\bfit{skew fibration} if for all $v \in \vG$ and $w\phi(v) \in \eH$
there exists $\tilde{w} \in \vG$ such that $\tilde{w}v \in \eG$ and
$\phi(\tilde{w})w \notin \eH$.  A directed graph homomorphism is a
\bfit{fibration} if for all $v \in \vG$ and $\pair{w,\phi(v)} \in
\eH$, there exists a unique $\tilde{w} \in \vG$ such that $\pair{\tilde{w},v}
\in \eG$ and $\phi(\tilde{w}) = w$.

A \bfit{fograph homomorphism} $\phi=\tuple{\phi,\substof\phi}$ is a
pair where $\phi\colon\gG\to\gH$ is a graph homomorphism between the
underlying graphs, and $\substof\phi$ is a variable renaming such that
for all $v\in\vG$ we have
$\labelof{\phi(v)}=\substof\phi(\labelof v)$. Note that this entails that
$\phi$ maps binders to binders and literals to literals. We say that a
fograph homomorphism $\phi\colon\gG \rightarrow \gH$ is
%%\bfit{label-preserving} if for all $v \in \vG, label(v) = label(f(v))$, and is
\bfit{existential-preserving} if for all existential binders $b$ in
$\gG$, the vertex $\phi(b)$ is an existential binder in $\gH$.

\begin{definition}
  Let $\gG$ and $\gH$ be fographs. A \bfit{skew bifibration}
  $\phi\colon\gG\to\gH$ is an existential-preserving fograph
  homomorphism that is a skew fibration on $\tuple{\vG, \eG} \to
  \tuple{\vH, \eH}$ and a fibration on the binding graphs $\bG\to\bH$.
  %etween fographs is a label-preserving and
%existential-preserving fograph homomorphism whose restriction to the
%binding graphs is a fibration.
\end{definition}

@@

\begin{remark}
  In or definition of FOCP, we are slightly laxer than the original
  definition of~\cite{hughes:fopws}, as we allow for a variable
  renaming $\substof\phi$ which was forced to be the identity
  in~\cite{hughes:fopws}.
  %The reason for this relaxation in our case is that we want to cover a richer class of proofs, also allowing 
\end{remark}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
